Classless Java: Tuning Java Interfaces

### 2015-11-27 paper/sections/CaseStudy.tex
Re read before korea completed

The \emph{Expression Problem} (EP)~\cite{wadler98expression}
is a well-known problem about modular extensibility issues in software
evolution. Recently, a new solution~\cite{eptrivially} that uses only
covariant type refinement that has been shown to work in Java. When
that solution is modelled with interfaces and default methods, it can
even provide independent extensibility: that is, the ability to
assemble a system, from multiple, independently developed extensions.
Unfortunatelly, a plain Java solution has alot of code verbosity in
Java due to code required for instantiation, which makes the solution 
cumbersome to use. The \mixin annotation helps to remove such code, and 
makes the solution quite easy to use.
% \bruno{We should try to do some space saving in here. You could, for example 
% combine the code for the initial system with the code for a new operation in a figure/table 
% with 2 colums. Then just refer to the Table/Figure in the text.}

@@

\begin{lstlisting}[multicols=2]
interface Exp { int eval(); }
@Obj interface Lit extends Exp {
	int x();
	default int eval() { return x(); }
}
@Obj interface Add extends Exp {
	Exp e1(); Exp e2();
	default int eval() {
		return e1().eval() + e2().eval();
	}
}
\end{lstlisting}

@@

\paragraph{Adding a New Type of Expressions}
It is easy to add new types of expressions to the code in the initial system in a
modular way. For example, the following code shows how to add subtraction.

@@

\paragraph{Adding a New Operation} The difficulty of the EP in object-oriented 
languages arizes from adding new operations. For example, adding a pretty printing 
operation, would typically involve changing all of the existing code. However, a solution 
to the EP forbids this and it also forbids using casts. In other words, it should be possible 
to add the pretty printing operation in a type-safe and modular way.  
It turns out that, this can be done easily, using our object interfaces approach. 
The following code shows how to add the new operation \texttt{print}.

\lstinputlisting[linerange=53-61]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
%\bruno{Please add a Java comment to the code emphasizing that the types of the getters are refined!}
%\bruno{please indent code properly (the getters are not aligned with default))}

>>>>

The \emph{Expression Problem} (EP)~\cite{wadler98expression}
is a well-known problem about modular extensibility issues in software
evolution. Recently, a new solution~\cite{eptrivially}, \marco{repharesed} using only
covariant type refinement, has been shown.
When such solution is modelled with interfaces and default methods, it can
even provide independent extensibility: the ability to
assemble a system from multiple, independently developed, extensions.
Unfortunatelly, the required instantiation code makes a plain Java solution verbose and cumbersome to use. The \mixin annotation is enough to remove the boilerplate code, making the presented approach very appealing.
% \bruno{We should try to do some space saving in here. You could, for example 
% combine the code for the initial system with the code for a new operation in a figure/table 
% with 2 colums. Then just refer to the Table/Figure in the text.}

@@

\begin{lstlisting}[multicols=2]
interface Exp { int eval(); }

@Obj interface Lit extends Exp {
	int x();
	default int eval() { return x(); }
}
@Obj interface Add extends Exp {
	Exp e1(); Exp e2();
	default int eval() {
		return e1().eval() + e2().eval();
	}
}
\end{lstlisting}

@@

\paragraph{Adding a New Type of Expressions}
In the object oriented paradigm, it is easy to add new types of expressions. For example, the following code shows how to add subtraction.

@@

\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in object-oriented 
languages arises from adding new operations. For example, adding a pretty printing 
operation, would typically involve changing all of the existing code. However, a solution 
to the EP forbids this and it also forbids using casts. In other words, it should be possible 
to add the pretty printing operation in a type-safe and modular way.  
It turns out that, this can be done easily, using our object interfaces approach. 
The following code shows how to add the new operation \texttt{print}.

\marco{ use the twoc colums here too}
\lstinputlisting[linerange=53-61]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
%\bruno{Please add a Java comment to the code emphasizing that the types of the getters are refined!}
%\bruno{please indent code properly (the getters are not aligned with default))}

### 2015-11-27 paper/sections/Formalization.tex
Re read before korea completed

During our experimentation, we found a bug in ECJ (eclipse version of javac): The case \textbf{B3},\textbf{A1}
is accepted by ECJ4.5.1 and rejected by javac.
By email communication with Brian Goetz (lead Java8 designer) we have confirmation that the expected behaviour is rejection, and that this is a bug in ECJ.
%Table~\ref{table:javabug} shows the method overriding bug in Java. In the
%example, there are 6 interfaces \texttt{A1,A2,A3,B1,B2,B3} with methods all
%named as \texttt{m} which are either abstract or default methods. If we define a
%new interface \texttt{M} that extends two of these interfaces, then the method
%overriding result is shown in the table. For example, row2 col2 means \texttt{M
%  extends A1,B1}. The result is \texttt{ERROR} because the abstract method
%\texttt{T m();} in \texttt{A1} conflicts with the method \texttt{default T m()
%  {...}} in \texttt{B1}. Readers may also figure out other extending cases by
%following this way of interpretation.

>>>>

During our experimentation, we found a bug in ECJ (eclipse version of javac): The case \textbf{B3},\textbf{A1}
is accepted by ECJ4.5.1 and rejected by javac.
By email communication with Brian Goetz (lead Java8 designer) we have confirmation that the expected behaviour is rejection, and that this is a bug in ECJ.\footnote{This bug is fixed in the developer version ECJ???, not yet released as stable.}
%Table~\ref{table:javabug} shows the method overriding bug in Java. In the
%example, there are 6 interfaces \texttt{A1,A2,A3,B1,B2,B3} with methods all
%named as \texttt{m} which are either abstract or default methods. If we define a
%new interface \texttt{M} that extends two of these interfaces, then the method
%overriding result is shown in the table. For example, row2 col2 means \texttt{M
%  extends A1,B1}. The result is \texttt{ERROR} because the abstract method
%\texttt{T m();} in \texttt{A1} conflicts with the method \texttt{default T m()
%  {...}} in \texttt{B1}. Readers may also figure out other extending cases by
%following this way of interpretation.

### 2015-11-27 paper/sections/FutureWork.tex
Re read before korea completed

\subsection{Qualifiers in Methods} %{Private state
The biggest limitation of our approach is the inability to model qualifiers
for class methods (private, protected, synchronized, etc.). For example, the absence
of the support for private/protected methods in Java8 interfaces forces all
members of interfaces to be public, including static methods. Since we use
abstract methods to encode the state, our state is always all public. Still, because 
the state can only be accessed by methods, it 
is impossible for the user to know if a certain method maps directly to a field
or if it has a default implementation.  If the user wants a constructor that
does not directly maps to the fields, (as for secondary constructors in Scala)
he can simply define its own \Q@of@ method and delegate on the generated one, as
in
\begin{lstlisting}
@Mixin interface Point{
  int x(); int y();
  static Point of(int val){return Point.of(val,val);}  
}
\end{lstlisting}
However, the generated \Q@of@ method would also be present and public.  If a
future version of Java was to support \emph{static private methods in
  interfaces} we could extend our code generation to handle also encapsulation.
Currently, it is possible to use a public nested class with private static
methods inside, but this is ugly and cumbersome. One possibility is that the
annotation processor takes code with a \Q!@Private! annotation, and turns it into
static private methods of a nested class. In this extension, also the \Q@of@
method could be made private following the same pattern.

\subsection{State initialization}
As discussed before, the user can trivially define its own \Q@of@ method, and
initialize a portion of the state with default values.  However, the
initialization code would not be reusable, and subinterfaces would have
to repeat such initialization code.  If a field has no setters, a simple
alternative is to just define the ``field'' as a default method as in
\begin{lstlisting}
@Mixin interface Box{ default int val(){return 0;} }
\end{lstlisting}
If setters are required, a possible extension of our code expansion could
recognize a field if the getter is provided and the setter is required, and
could generate the following code:
\lstinputlisting[linerange=4-11]{../UseMixinLombok/src/test/TestStateInitialization.java}% APPLY:linerange=STATE_INIT

We are unsure about the value of this workaround because of its trickiness. In
order to define a state with initialization, users have to learn this pattern:
defining a method which will be overridden later with behaviour change, but
change only after the first setter is called. Moreover, this code would cache
the result instead of re-computing it every time. This can be very tricky in a
non functional setting.
%\bruno{Yanlin please polish text/break long sentences.}

@@

Methods originally defined in Java class \Q@Object@, as \Q@clone@ and
\Q@toString@, can be supported by our approach, but they need special
care. If an interface annotated with \mixin asks an implementation for
\Q@clone@, \Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily
generate one from the fields.\footnote{In particular, for clone we can
  do automatic return type refinement as we do for \Q@with-@ and
  fluent setters. Note how this would solve most of the Java ugliness
  related to \Q@clone@ methods.}
However, if the user wishes to provide his own implementation, since the method is
also implemented in \Q@Object@, a conflict arises. The user has to
explicitly resolve the conflict inside \Q@of@, by implementing the method and
delegating it to the user implementation, thus
\begin{lstlisting}
@Mixin interface Point{ int x(); int y();
  default Point clone(){ return Point.of(0,0);}//user defined clone
}
\end{lstlisting} 
Would expand into

>>>>

\subsection{Qualifiers in Methods} %{Private state
The biggest limitation of our approach is the inability to model qualifiers
for class methods (private, protected, synchronized, etc.). For example, the absence
of the support for private/protected methods in Java8 interfaces forces all
members of interfaces to be public, including static methods. Since we use
abstract methods to encode the state, our state is always all public. Still, because 
the state can only be accessed by methods, it 
is impossible for the user to know if a certain method maps directly to a field
or if it has a default implementation.  If the user wants a constructor that
does not directly maps to the fields, (as for secondary constructors in Scala)
he can simply define its own \Q@of@ method and delegate on the generated one, as
in
\begin{lstlisting}
@Obj interface Point{  int x(); int y();
  static Point of(int val){return Point.of(val,val);}  }
\end{lstlisting}
However, the generated \Q@of@ method would also be present and public.  If a
future version of Java was to support \emph{static private methods in
  interfaces} we could extend our code generation to handle also encapsulation.
Currently, it is possible to use a public nested class with private static
methods inside, but this is ugly and cumbersome. One possibility is that the
annotation processor takes methods with a \Q!@Private! annotation, and turns it into
static private methods of a nested class. In this extension, also the \Q@of@
method could be made private following the same pattern.

\subsection{State initialization}
As discussed before, the user can trivially define its own \Q@of@ method, and
initialize a portion of the state with default values.  However, the
initialization code would not be reusable, and subinterfaces would have
to repeat such initialization code.  If a field has no setters, a simple
alternative is to just define the ``field'' as a default method as in
\begin{lstlisting}
@Obj interface Box{ default int val(){return 0;} }
\end{lstlisting}
If setters are required, a possible extension of our code expansion could
recognize a field if the getter is provided and the setter is required, and
could generate the following code:
\lstinputlisting[linerange=4-11]{../UseMixinLombok/src/test/TestStateInitialization.java}% APPLY:linerange=STATE_INIT

We are unsure about the value of this workaround because of its trickiness. In
order to define a state with initialization, users have to defining a method trusting that it will be overridden later with a behavioural change, but
such change is visible only after the first time the setter is called. Moreover, this code would cache
the result instead of re-computing it every time. This can be very tricky in a
non functional setting.
%\bruno{Yanlin please polish text/break long sentences.}

@@

Methods originally defined in Java class \Q@Object@, as \Q@clone@ and
\Q@toString@, can be supported by our approach, but they need special
care. If an interface annotated with \mixin asks an implementation for
\Q@clone@, \Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily
generate one from the fields.\footnote{In particular, for clone we can
  do automatic return type refinement as we do for \Q@with-@ and
  fluent setters. Note how this would solve most of the Java ugliness
  related to \Q@clone@ methods.}
However, if the user wishes to provide his own implementation, since the method is
also implemented in \Q@Object@, a conflict arises. The generated code can resolve the conflict inside \Q@of@, by implementing the method and
delegating it to the user implementation, thus
\begin{lstlisting}
@Obj interface Point{ int x(); int y();
  default Point clone(){ return Point.of(0,0);}//user defined clone
}
\end{lstlisting} 
Would expand into

### 2015-11-27 paper/sections/Overview.tex
Re read before korea completed

\begin{lstlisting}
@Mixin interface Point2D{ int x(); int y();}
\end{lstlisting}

@@

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new piece of data by reusing most
information from an old piece of data is very common
when programming with immutable data-structures; it is so common that
it is supported in our code generation as \Q@with@ methods,
that is:
\begin{lstlisting}
@Mixin interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods
}
\end{lstlisting}

@@

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, they
just provide it using a \Q@default@ method. For example:
\begin{lstlisting}
@Mixin interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

@@

\paragraph{\Q@Point3D@ and properties updater}
To implement the old setter in a convenient way, \mixin supports one
last type of operations: property updater \texttt{with}
methods. Unlike the \texttt{withX} (where \texttt{X} stands for a
field name) methods presented so far, property updaters take several
fields at once, contained in an interface, and copy those fields into
fields of another interface.
The \Q@Point3D@ interface is defined as follows:
\lstinputlisting[linerange=51-55]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\noindent By using the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store.  Note how all the informations about what fields sits in
\Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent for the implementer of \Q@Bird@.

>>>>

\begin{lstlisting}
@Obj interface Point2D{ int x(); int y();}
\end{lstlisting}

@@

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new piece of data by reusing most
information from an old piece of data is very common
when programming with immutable data-structures; it is so common that
it is supported in our code generation as \Q@with@ methods,
that is:
\begin{lstlisting}
@Obj interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods
}
\end{lstlisting}

@@

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, they
just provide it using a \Q@default@ method. For example:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

@@

\paragraph{\Q@Point3D@ and properties updater}
To implement the old setter in a convenient way, \mixin supports one
last type of operations: property updater \texttt{with}
methods. Unlike the \texttt{withX} (where \texttt{X} stands for a
field name) methods presented so far, property updaters take several
fields at once, contained in an interface, and copy those fields into
fields of another interface.
The \Q@Point3D@ interface is defined as follows:
\lstinputlisting[linerange=51-56]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\noindent By using the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store.  Note how all the informations about what fields sits in
\Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent for the implementer of \Q@Bird@.

### 2015-11-27 paper/sections/RelatedWork.tex
Re read before korea completed

Here compile time code generation comes at the rescue: if for a certain feature
(\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature (annotating a
class and provide getters), then we can insert in the compilation process a tool
that examins and enrichs the code before compilation. No need to modify the
original source; for example we can work on temporary files Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process: in this way
there is no need of temporary files, and there is a well defined way of putting
multiple extensions together.

@@

\subsection{Formalization of Java8}
We provide a simple formalization for a subset of Java including default/static
interface methods and object initialization literals (often called anonymous
local inner classes).  A similar formalization was drafted by Goetz and
Field~\cite{goetz12fdefenders} to formalize defender (default) methods in
Java. However this formalization is limited to model exactly one method inside
classes/interfaces.\bruno{explain why having multiple methods is significantly
  more challenging.} Our code generation generates several methods that may
interacts with each other and methods defined by user. Thus implementing and
formalizing multiple methods inside interfaces adds complexity greatly.

% As an evidence of the attention and care present in our formalization work,
% while double-checking the behaviour of Java in side cases we have discovered a
% likely bug in the current \texttt{javac}.\marco{refer to before when we explain
%   the issue.}

>>>>

Here compile time code generation comes at the rescue: if for a certain feature
(\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examins and enrichs the code before compilation. No need to modify the
original source; for example we can work on temporary files. Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process: in this way
there is no need of temporary files, and there is a well defined way of putting
multiple extensions together.

@@

\subsection{Formalization of Java8}
We provide a simple formalization for a subset of Java including default/static
interface methods and object initialization literals (often called anonymous
local inner classes).  A similar formalization was drafted by Goetz and
Field~\cite{goetz12fdefenders} to formalize defender (default) methods in
Java. However this formalization is limited to model exactly one method inside
classes/interfaces.\bruno{explain why having multiple methods is significantly
  more challenging.} Our code generation generates several methods that may
interacts with each other and methods defined by user. Thus implementing and
formalizing multiple methods inside interfaces adds complexity greatly.
\marco{I do not see this as a good explanation, I think we may just remove the `one method limitation discussion'. Is their work modelling interactions between abstract and default methods? like making a default method abstract again?}
% As an evidence of the attention and care present in our formalization work,
% while double-checking the behaviour of Java in side cases we have discovered a
% likely bug in the current \texttt{javac}.\marco{refer to before when we explain
%   the issue.}

### 2015-11-27 paper/sections/Translation.tex
Re read before korea completed

\subsection{Results}
\textbf{THEOREM. }
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK,\\*
if $\II_0$ has \mixin,
$\valid(\C_0)$  and $\QM{of}\notin\dom(\C_0)$,
there is no $\C<:\C_0$, there is no \Q@new @$\C_0$\Q@(){..}@,
then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$\\*
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ $\II$ OK.

@@

\begin{lstlisting}
interface A { int x(); A withX(int x); }
@Mixin interface B extends A {}
interface C extends B { A withX(int x); }
\end{lstlisting}

>>>>

\subsection{Results}
Informally, in a type correct interface table, if we translate a specific interface, then
the resulting table is type correct, or there is a subtype of the translated interface in the original table.
\textbf{THEOREM. }
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK,\\*
if $\II_0$ has \mixin,
$\valid(\C_0)$  and $\QM{of}\notin\dom(\C_0)$,
there is no $\C<:\C_0$, there is no \Q@new @$\C_0$\Q@(){..}@,
then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$\\*
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ $\II$ OK.

@@

\begin{lstlisting}
interface A { int x(); A withX(int x); }
@Obj interface B extends A {}
interface C extends B { A withX(int x); }
\end{lstlisting}
